---
title: "Using vacsce for vaccination coverage scenario generation"
author: "Xiang Li"
date: "`r Sys.Date()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using vacsce for vaccination coverage scenario generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library("dplyr")
#library(vacsce)
devtools::load_all("../")
```

### Introduction

This vignette describes how to use *vacsce* to generate vaccination coverage scenarios. *vacsce* provides a set of coverage projection rules. Its initial objective was to document commonly used coverage projection rules and works as a tool for VIMC's model runs, interim-update and related analysis. As VIMC continues expanding its collaboration with greater community, we further plan to release *vacsce* as a tool for external users who share common interests such as planning for immunization programs, assessment of vaccine impact, etc.

### Future vaccination coverage projection rules

*vacsce* currently contains seven coverage projection rules: four of which are related to routine immunization and three for supplementary immunization activities (SIAs) (see *vacsce/R/projection_rules.R*). In this section, we provide details of these rules with examples.

**Load example data**
```{r}
data <- read.csv("inst/")
year_cur <- 2021
his <- data[data$year <= year_cur & data$year > 2000, ]
fut <- data[data$year > year_cur, ]
```

**keep_levels**
Routine coverage projection - Future coverage level is maintained at certain constant level.

```{r}
d <- his[his$vaccine == "MCV1",]
keep_levels(d, year_from = year_cur + 1, year_to=2030, level=d$coverage[d$year == year_cur])
```
In this example, the historical coverage of MCV1 for country ISO is `r d$coverage[d$year == year_cur]` in year `r year_cur`. *keep_levels* projects future coverage from `r year_cur+1` to 2030, where future coverage level is maintained at the level of `r year_cur`.


**incremental**
Routine coverage projection - Future coverage increase incrementally.

```{r}
d <- his[his$vaccine == "MCV1",]
incremental(d, year_from = year_cur + 1, year_to = 2030, step = 0.01, cap = 0.95)
```
The historical coverage of MCV1 for country ISO is `r d$coverage[d$year == year_cur]` in year `r year_cur`. *incremental* projects future coverage from `r year_cur+1` to 2030, where future coverage level increases by 1% (set by parameter - *step*) annually. Parameter *cap* sets a threshold for maximum future coverage level. However, it will be coerced by best historical coverage is which is higher.

**catch_up_with_x**

```{r}
d <- his[his$vaccine == "MCV2",] # zero-rows as not historical data
catch_up_with_x(d, year_from = year_cur + 3, year_to = year_cur + 7, vaccine_x_level = 0.85, intro_level = 1/3) %>%
  mutate(disease = "Measles", vaccine = "MCV2", country = "ISO", activity_type = "routine", age_from = 2, age_to = 2, gender = 1, proportion_risk = 1)
```
There is no historical MCV2 routine introduction in the example data. *catch_up_with_x* projects routine introduction in year `r year_cur + 3`. Coverage level increases non-linearly and reaches 85% in year `r year_cur + 6`. 85% represents coverage level of a certain vaccine in year `r year_cur + 6`. This coverage level multiply by factor *intro_level* determines the routine introduction coverage level of MCV2 in `r year_cur + 3`. 

**non_linear_scale_up**

```{r}
d <- his[his$vaccine == "MCV1",]
non_linear_scale_up(d, year_from = year_cur + 1, year_to = 2030, endpoint = 0.9)
```
The historical coverage of MCV1 for country ISO is `r d$coverage[d$year == year_cur]` in year `r year_cur`. *non_linear_scale_up* projects future coverage from `r year_cur+1` to 2030, where future coverage level non-lineraly increases annually. Parameter *endpoint* sets the target coverage level in year 2030.

**sia_follow_up**

```{r}
d <- his[his$vaccine == "MCV1",]
dat <- non_linear_scale_up(d, year_from = year_cur + 1, year_to = 2030, endpoint = 0.9) %>%
  mutate(country = d$country[1], vaccine = d$vaccine[1])
d <- his[his$vaccine == "Measles", ]
sia_follow_up(d, dat, vaccine_base = "MCV1", year_current = year_cur, year_to = 2030, look_back = 4, sia_level = 0.9, age_from = 1, age_to = 5) %>%
   mutate(disease = "Measles", vaccine = "Measles", country = "ISO", activity_type = "campaign", gender = 1, proportion_risk = 1)
```
Projection rule, *sia_follow_up*, projects future recurrent follow-up supplementary immunisation activities (SIAs) according to a baseline routine vaccination trajectory and historical SIAs. In this example, we first project future MCV1 coverage using the *non_linear_scale_up* rule. Then we evaluate MCV1 coverage trajectory and historical SIAs from year *year_current-look_back* to determine future follow-up campaigns. For details of how this evaluation is done, see projection_rules.R/sia_follow_up for details.

**sia_catch_up**
```{r}
d <- his[his$vaccine == "HPV",]
dat <- catch_up_with_x(d, year_from = 2027, year_to = 2030, vaccine_x_level = 0.7) %>%
   mutate(disease = "HPV", vaccine = "HPV", country = "ISO", activity_type = "routine", age_from = 9, age_to = 9, gender = 3, proportion_risk = 1)
sia_catch_up(d, dat, vaccine_base = "HPV", year_current = year_cur, sia_level = 0.9, age_from = 10, age_to = 15)%>%
   mutate(disease = "HPV", vaccine = "HPV", country = "ISO", activity_type = "campaign", gender = 3, proportion_risk = 1)
```

```{r}
d <- his[his$vaccine == "MenA" & his$activity_type == "routine",]
dat <- catch_up_with_x(d, year_from = 2023, year_to = 2026, vaccine_x_level = 0.7) %>%
   mutate(disease = "MenA", vaccine = "MenA", country = "ISO", activity_type = "routine", age_from = 9, age_to = 9, gender = 3, proportion_risk = 1)
d <- his[his$vaccine == "MenA" & his$activity_type == "campaign",]
sia_catch_up(d, dat, vaccine_base = "MenA", year_current = year_cur, sia_level = 0.9, age_from = 1, age_to = 29)%>%
   mutate(disease = "MenA", vaccine = "MenA", country = "ISO", activity_type = "campaign", gender = 1, proportion_risk = 1)
```
**sia_recurrent**

```{r}
d <- his[his$vaccine == "Measles", ]
sia_recurrent(d, dat = NULL, year_from = 2020, year_to = 2030, frequency = 3, sia_level = 0.3, age_from = 1, age_to = 60)%>%
   mutate(disease = "Measles", vaccine = "Measles", country = "ISO", activity_type = "campaign", gender = 1, proportion_risk = 1)
```

```{r}
d <- his[his$vaccine == "Cholera", ]
sia_recurrent(d, dat = NULL, year_from = 2020, year_to = 2030, frequency = 3, sia_level = 0.3, age_from = 1, age_to = 60)%>%
   mutate(disease = "Cholera", vaccine = "Cholera", country = "ISO", activity_type = "campaign", gender = 1, proportion_risk = 1)
```
